---
title: "Integration of the second batch of CAGE libraries."
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
...

# R commands to collect the numbers (you can skip reading it.)

```{r echo=FALSE}
#knitr::opts_chunk$set(cache  = TRUE)
knitr::opts_knit$set(verbose = TRUE)
# knitr::opts_chunk$set(dev=c('svg', 'png'))
options(width = 100)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r load_libraries}
library(ggplot2)
library(dplyr)
library(magrittr)
```

```{r load_sample_metadata}
samples   <- read.delim("../generated/samples.tsv", head = TRUE, sep = "\t")
libraries <- read.table('../source/libraries.tsv'
                       , header = TRUE
                       , stringsAsFactors = FALSE
                       , colClasses = c(Moirai_Stamp="character"))
```

```{r useful_aliases}
round2 <- libraries[grep("CAGE_[456]", libraries$Name), "C1_Run_ID"]
```

```{r integrate-per-lib-qc}
# # This only works when the cached results are available.
# 
# cage4 <- new.env()
# local(rmarkdown::render("RNhi10538.Rmd"), cage4)
# 
# cage5 <- new.env()
# local(rmarkdown::render("RNhi10539.Rmd"), cage5)
# 
# cage6 <- new.env()
# local(rmarkdown::render("RNhi10540.Rmd"), cage6)
# 
# libs <- rbind(cage4$libs, cage5$libs, cage6$libs)
# libs$Rownames <- libs$samplename

libs <- rbind(
  read.table("RNhi10538.libs.tsv", sep="\t", head=TRUE),
  read.table("RNhi10539.libs.tsv", sep="\t", head=TRUE),
  read.table("RNhi10540.libs.tsv", sep="\t", head=TRUE)
)

libs$Rownames <- libs$samplename

samples %<>% left_join(libs[,c("Rownames", "l1", "r100l1", "genes", "geneSymbols")])

samples[samples$Run %in% round2, ] %>% summary
samples[samples$Run %in% round2, c("Run","Timepoint")] %>% gdata::drop.levels() %>% table
samples[samples$Run %in% round2, c("Final","Keep")] %>% gdata::drop.levels() %>% table

samples[                         , "Group"] <- samples$Final %>% as.character
samples[!is.na(samples$Timepoint), "Group"] <- samples[!is.na(samples$Timepoint), "Timepoint"] %>% as.character
samples[!is.na(samples$Control),   "Group"] <- samples[!is.na(samples$Control), "Control"] %>% as.character
samples$Group %<>% factor

samples$Experiment <- sub("_...$", "", samples$Rownames)
```

# Quick summaries

```{r round2_summary}
myPlot <- function(X) {
  set.seed(1)
  X <- substitute(X)
  ggplot( samples[samples$Run %in% round2,] %>%
            gdata::drop.levels() %>%
            sample_frac) +
    coord_flip() +
    geom_jitter( width = 0.3
               , aes( color=Experiment
                    , Group
                    , eval(X))) +
    ylab(deparse(X))
}
```

## Indicators that scale with depth.

Values for CAGE_5 should be lower, since it had only one HiSeq lane.

### Number of genes detected.

```{r summary-genes}
myPlot(genes)
```

### Number of L1 TSS detected.

```{r summary-l1}
myPlot(l1)
```

### Number of molecules detected.

```{r summary-counts}
myPlot(counts)
```

## Indicators that should not vary with depth

### Richness (scale of 100)

```{r summary-richness}
myPlot(r100l1)
```

### cDNA concentration

CAGE_4 run looks cleaner.

```{r summary-concentration}
myPlot(Concentration)
```

### Spikes

According to empty chambers, the amount of spikes varied between runs.
(`CAGE_4` > `CAGE_5` > `CAGE_6`).

```{r summary-spikes}
myPlot(spikes / total)
```

### rRNA

Empty chambers indicate background values (environmental
contaminations ?).  Negative control shows that spike RNA
does not result in significant amounts of artificial rRNA
matches.

More rRNA is observed in `CAGE_6` when there is a cell, but
this can be just a consequence of having less spikes (see
above).

```{r summary-rRNA}
myPlot(rdna / total)
```

### rRNA / spikes ratio

Higher in case of the presence of a cell.

```{r summary-rnda-spikes}
myPlot(rdna / spikes)
```

### Tag dust

The `CAGE_5` run is much nosier.

```{r summary-tagdust}
myPlot(tagdust / total)
```

